Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-06-01 iso-int/rc-iso-int.tex
small edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% Previous work on {{langRevT}} \cite{rc2011,infeffects} studied its
% properties and various constructions in it. However, for a
% computational model, there has been little focus on actually writing
% programs directly in {{langRevT}}. This paper aims to remedy that by
% presenting a systematic technique of expressing logically reversible
% small-step abstract machines in {{langRevT}}. In other words, this
% paper shows that once we have a logically reversible machine in a
% notation of our choice, expressing the machine as an isomorphic
% interpreter in {{langRevT}} can be done systematically and does not
% present any significant conceptual difficulties. This gives us a means
% of developing large {{langRevT}} programs with the ease of reasoning
% at the level of a conventional small-step semantics.  We develop
% several simple interpreters over numbers and addition, move onto tree
% traversal and finish with an interpreter for {{langRevT}} expressed in
% {{langRevT}}.
  In our previous work, we developed a reversible programming language and
  established that every computation in it is a (partial) isomorphism that is
  reversible and that preserves information. The language is founded on type
  isomorphisms that have a clear categorical semantics but that are awkward
  as a notation for writing actual programs, especially recursive ones. This
  paper remedies this aspect by presenting a systematic technique for
  developing a large and expressive class of reversible recursive programs,
  that of logically reversible small-step abstract machines. In other words,
  this paper shows that once we have a logically reversible machine in a
  notation of our choice, expressing the machine as an isomorphic interpreter
  can be done systematically and does not present any significant conceptual
  difficulties. Concretely, we develop several simple interpreters over
  numbers and addition, move on to tree traversals, and finish with a
  meta-circular interpreter for our reversible language. Generally, this
  gives us a means of developing large reversible programs with the ease of
  reasoning at the level of a conventional small-step semantics.
\end{abstract}

@@

In this paper, we investigate the practicality of writing recursive programs
in {{langRevT}}. Specifically, we assume that we are given some reversible
recursive algorithm expressed as a \emph{reversible abstract machine}, and we
show via a number of systematic steps, how to develop a {{langRevT}} program
from that abstract machine. Our choice of starting from reversible abstract
machines is supported by the following observations:
\begin{itemize}
\item it is an interesting enough class of reversible programs: researchers
  have invested the effort in manually designing such machines, e.g., the
  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and the SECD-H~\cite{lorenz};
\item every reversible interpreter can be realized as such a machine: this
  means that our class of programs includes self-interpreters which are
  arguably a measure of the strength of any reversible
  language~\cite{Axelsen:2011:RPC:1987171.1987176,Yokoyama:2008:PRP,Yokoyama:2007:RPL:1244381.1244404};
\item general recursive programs can be systematically transformed to
  abstract machines: the technique is independent of reversible programs and
  consists of transforming general recursion to tail recursion and then
  applying \emph{fission} to split the program into a driver and a small-step
  machine~\cite{Danvy:2008:RRN:1813347.1813350,Rendel:2010:ISD:1863523.1863525}.
\end{itemize}

@@

\item Associativity is implicit in the graphical language. Three parallel
  wires may represent {{b1*(b2*b3)}} or {{(b1*b2)*b3}}.
%% \begin{center}
%% \scalebox{0.95}{
%% %%subcode-line{pdfimage}[diagrams/thesis/associate.pdf]
%% \includegraphics{diagrams/thesis/assoc.pdf}
%% }
%% \end{center}

@@

\vspace{-20px}
% \begin{multicols}{2}
\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/nat-nat6.pdf}
}
\end{center}

@@

%\columnbreak
\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/nat-nat7.pdf}
}
\end{center}  
%\end{multicols}

%% The final figure is given below:

@@

%% \begin{center}
%% \scalebox{1.0}{ \includegraphics{diagrams/nat-nat8.pdf} }
%% \end{center}

@@

Although trivial, the previous example captures the fundamental aspects of
our construction. In general, our starting point is an abstract machine in
which every rewrite step is logically reversible. The formal criterion of
logical reversibility in this setting is captured by Abramsky's
\emph{bi-orthogonal automata}~\cite[Sec.~2]{abramsky2005structural} which is
summarized as:
\begin{enumerate}
\item
  Machines must have distinguishable start and stop states. There may
  be more than one valid start state and more than one valid stop
  state, however.
\item Each valid machine state must match a unique pattern on the
  left-hand side and right-hand side of the `{{|-->}}' relation.
\item Every reduction step must be computable (in {{langRevT}}) and
  must be logically reversible --- i.e. it must be possible to reduce
  from right to left.
\end{enumerate}
Given such a machine the process of encoding the machine in {{langRevT}}
consists of the following steps:
\begin{enumerate}
\item Expand the input to expose enough structure to distinguish the
  left-hand side of each machine state;
\item Expand the output to expose enough structure to distinguish the
  right-hand side of each machine state;
\item Shuffle matching input terms to output terms, inserting any appropriate
  mediating computations. 
\item Slide the two sections to expose the start and stop state and
  introduce a {{trace}} to iterate the construction. 
\end{enumerate}

@@

The reduction rules on the right traverse a given tree and increment
every leaf value.  The machine here is a little richer than the ones
dealt with previously. In particular, we have a two types of machine
states {{<t,c>}} and {{ {[t, c]} }}. The first of these corresponds
to walking down a tree, building up the context in the process. The
second corresponds to reconstructing the tree from the context and
also switching focus to any unexplored subtrees in the process. There
are also two syntactic categories to deal with (trees and tree
contexts) where previously we only had numbers. The {{fold}} and
{{unfold}} isomorphisms that we need for trees and tree contexts are:

@@

%subcode{bnf} include main
%! columnStyle = rrcll
% unfold_t :& ~~~~t &<->& n + t * t &: fold_t
% unfold_c :& c &<->& 1 + c * t + t * c ~~~~~~&: fold_c
\end{small}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% Previous work on {{langRevT}} \cite{rc2011,infeffects} studied its
% properties and various constructions in it. However, for a
% computational model, there has been little focus on actually writing
% programs directly in {{langRevT}}. This paper aims to remedy that by
% presenting a systematic technique of expressing logically reversible
% small-step abstract machines in {{langRevT}}. In other words, this
% paper shows that once we have a logically reversible machine in a
% notation of our choice, expressing the machine as an isomorphic
% interpreter in {{langRevT}} can be done systematically and does not
% present any significant conceptual difficulties. This gives us a means
% of developing large {{langRevT}} programs with the ease of reasoning
% at the level of a conventional small-step semantics.  We develop
% several simple interpreters over numbers and addition, move onto tree
% traversal and finish with an interpreter for {{langRevT}} expressed in
% {{langRevT}}.
  In our previous work, we developed a reversible programming language and
  established that every computation in it is a (partial) isomorphism that is
  reversible and that preserves information. The language is founded on type
  isomorphisms that have a clear categorical semantics but that are awkward
  as a notation for writing actual programs, especially recursive ones. This
  paper remedies this aspect by presenting a systematic technique for
  developing a large and expressive class of reversible recursive programs,
  that of logically reversible small-step abstract machines. In other words,
  this paper shows that once we have a logically reversible machine in a
  notation of our choice, expressing the machine as an isomorphic interpreter
  can be done systematically and does not present any significant conceptual
  difficulties. Concretely, we develop several simple interpreters over
  numbers and addition, move on to tree traversals, and finish with a
  meta-circular interpreter for our reversible language. This
  gives us a means of developing large reversible programs with the ease of
  reasoning at the level of a conventional small-step semantics.
\end{abstract}

@@

In this paper, we investigate the practicality of writing large recursive programs
in {{langRevT}}. Specifically, we assume that we are given some reversible
recursive algorithm expressed as a \emph{reversible abstract machine}, and we
show via a number of systematic steps, how to develop a {{langRevT}} program
from that abstract machine. Our choice of starting from reversible abstract
machines is supported by the following observations:
\begin{itemize}
\item it is an interesting enough class of reversible programs: researchers
  have invested the effort in manually designing such machines, e.g., the
  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and the SECD-H~\cite{lorenz};
\item every reversible interpreter can be realized as such a machine: this
  means that our class of programs includes self-interpreters which are
  arguably a measure of the strength of any reversible
  language~\cite{Axelsen:2011:RPC:1987171.1987176,Yokoyama:2008:PRP,Yokoyama:2007:RPL:1244381.1244404};
\item general recursive programs can be systematically transformed to
  abstract machines: the technique is independent of reversible programs and
  consists of transforming general recursion to tail recursion and then
  applying \emph{fission} to split the program into a driver and a small-step
  machine~\cite{Danvy:2008:RRN:1813347.1813350,Rendel:2010:ISD:1863523.1863525}.
\end{itemize}

@@

\item Associativity is implicit in the graphical language. Thus, three parallel
  wires may represent {{b1*(b2*b3)}} or {{(b1*b2)*b3}}.
%% \begin{center}
%% \scalebox{0.95}{
%% %%subcode-line{pdfimage}[diagrams/thesis/associate.pdf]
%% \includegraphics{diagrams/thesis/assoc.pdf}
%% }
%% \end{center}

@@

\vspace{-20px}
% \begin{multicols}{2}
\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/nat-nat6.pdf}
}
\end{center}

@@

%\columnbreak
\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/nat-nat7.pdf}
}
\end{center}  
%\end{multicols}

\noindent
The completed interpreter is given below:

@@

\begin{center}
\scalebox{1.0}{ \includegraphics{diagrams/nat-nat8.pdf} }
\end{center}

@@

Although trivial, the previous example captures the fundamental aspects of
our construction. In general, our starting point is an abstract machine in
which every rewrite step is logically reversible. The formal criterion of
logical reversibility in this setting is captured by Abramsky's
\emph{bi-orthogonal automata}~\cite[Sec.~2]{abramsky2005structural} which is
summarized as:
\begin{enumerate}
\item
  Machines must have distinguishable start and stop states. There may
  be more than one valid start state and more than one valid stop
  state, however.
\item Each valid machine state must match a unique pattern on the
  left-hand side and right-hand side of the `{{|-->}}' relation.
\item Every reduction step must be computable (in {{langRevT}}), must
  not drop or duplicate pattern variables and must be logically
  reversible --- i.e. it must be possible to reduce from right to
  left.
\end{enumerate}
Given such a machine the process of encoding the machine in {{langRevT}}
consists of the following steps:
\begin{enumerate}
\item Expand the input to expose enough structure to distinguish the
  left-hand side of each machine state;
\item Expand the output to expose enough structure to distinguish the
  right-hand side of each machine state;
\item Shuffle matching input terms to output terms, inserting any appropriate
  mediating computations. 
\item Slide the two sections to expose the start and stop state and
  introduce a {{trace}} to iterate the construction. 
\end{enumerate}

@@

The reduction rules on the right traverse a given tree and increment
every leaf value.  The machine here is a little richer than the ones
dealt with previously. In particular, we have two types of machine
states {{<t,c>}} and {{ {[t, c]} }}. The first of these corresponds
to walking down a tree, building up the context in the process. The
second corresponds to reconstructing the tree from the context and
also switching focus to any unexplored subtrees in the process. There
are also two syntactic categories to deal with (trees and tree
contexts) where previously we only had numbers. The {{fold}} and
{{unfold}} isomorphisms that we need for trees and tree contexts are:

@@

%subcode{bnf} include main
%! columnStyle = rcl
% unfold_t :& t <-> n + t * t &: fold_t
% unfold_c :& ~~~~~ c <-> 1 + c * t + t * c ~~~~ &: fold_c
\end{small}
